---
title: "How Rule Changes in Baseball Affected The Game in History"
author: "Faith Rhinehart"
date: "2026-05-4"
execute: 
  echo: true
  warning: false
format: 
  html:
    embed-resources: true
---

```{r}
library(readr)
library(ggplot2)
library(tidyverse)
library(dplyr)
mlb_teams <- read_csv("data-finalProject/mlb_teams.csv")
Baseball <- mlb_teams |> select(1:2, 13, 18, 21, 35) |> slice(261:2784)
```

**Abstract**

*Baseball has undergone numerous rule changes since its formalization in the mid‑19th century, many of which have fundamentally altered how the game is played and how player performance is measured. This project examines several major rule changes in Major League Baseball history and analyzes their impact on key offensive and pitching statistics. Using the mlb_teams dataset from OpenIntro which includes team‑level statistics from 1876 to 2020, this analysis focuses on three major rule changes: the abolition of the spitball, the introduction of the designated hitter, and changes to the strike zone. Through visualizations and comparative analysis, this project highlights how these rule changes influenced statistical trends such as runs scored, stolen bases, home runs, and strikeouts. Overall, the results show that rule changes have played a significant role in reshaping baseball strategy and performance across eras.*

**Introduction**

*Created in 1840s New York, modern baseball took on its first set of formal rules that began to shape and define the sport. These rules included the diamond‑shaped infield, three outs per inning, and foul lines. By 1876, baseball had grown in popularity and professional teams were introduced, eventually leading to the creation of the National League (NL) in 1876 and the American League (AL) in 1901, establishing baseball as the oldest professional sport in the United States.*

*Since the sport has now been played for over 150 years, there have been many rule changes. Whether introduced to improve the game or better define it, these rule changes made a lasting impact on how baseball is played. The goal of this project is to examine how some of the most significant rule changes in baseball history impacted overall team statistics.The mlb_teams dataset from OpenIntro (Linked Here: https://www.openintro.org/data/index.php?data=mlb_teams) contains information on every professional team that exists or has existed from 1876 to 2020. The dataset includes 41 variables with 2,784 observations. Key variables include year of play, league, division, win‑loss records, postseason results, offensive statistics (such as runs scored, hits, home runs, walks, strikeouts, and stolen bases), pitching statistics (such as earned runs, ERA, strikeouts, and shutouts), and fielding metrics. Additional metadata includes team name, ballpark, and home attendance.* 

*To examine the effects of rule changes, the dataset was sliced to begin at 1903. In that year, the first World Series was played, marking the beginning of baseball’s “Modern Era” and introducing postseason play, which influenced team statistics. Excluding years before 1903 helps control for this structural change. The variables of interest in this analysis are year, league_id, runs_scored, homeruns, stolen_bases, and strikeouts_by_pitchers. In 1961 (AL) and 1962 (NL), baseball set a standardized 162 games for the regular season which means years before that had varying game numbers. Also, NA values were removed from the dataset. Both of these should be considered when interpreting the results. The cleaned dataset is referred to as "Baseball".*


**Rule Change #1: The Abolition of Spitball**

*Before 1920, pitchers were allowed to apply foreign substances such as dirt or pine tar to the baseball in order to deceive batters or improve grip. This contributed to what is known as the “dead‑ball era,” which was characterized by low scoring, frequent stolen bases, and small‑ball strategies used to manufacture runs. In 1920, MLB banned the use of foreign substances on the baseball. Because baseballs were white, this rule was relatively easy to enforce and also made the ball easier for batters to see.*

*This change shifted the game in favor of hitters and significantly altered offensive strategy. As batters began hitting the ball farther, teams relied less on stolen bases and small‑ball tactics. As a result, stolen bases declined while runs scored increased after 1920. Prior to the rule change, advancing runners through hitting was difficult because most balls stayed in the infield, making outs more likely. After 1920, harder and longer hits allowed runners to advance more easily without the need to steal bases.*

*The following graphs highlight how stolen bases and runs scored changed over time. The year range is restricted to 1903–1940 to emphasize the era most affected by this rule change. Changes occurring after 1940 could be influenced by additional factors beyond the abolition of the spitball.*

```{r}
stolen_bases <- Baseball |> filter(year <= "1940") |>
mutate(period = if_else(year < 1920, "pre-1920", "post-1920"))

ggplot(data = stolen_bases, aes(x = year, y = stolen_bases, colour = period)) +
  geom_point() +
  geom_smooth(color = "black") +
  labs(title = "Stolen Bases Through The Years 1903 - 1940",
       x = "Year",
       y = "Number of Stolen Bases",
       colour = "Time Period") +
  scale_color_brewer(palette = "Set2") +
  theme_minimal()
```

*In the graph, the number of stolen bases begins to decline during the 1910s. While this decrease started prior to 1920, the trend line shows that stolen bases continued to decline after the rule change and eventually reached their lowest point. The post‑1920 observations consistently show fewer stolen bases than the pre‑1920 observations, supporting the conclusion that the rule change impacted base‑stealing strategy.*

```{r}
runs_scored <- Baseball |> filter(year <= "1940") |>
  mutate(period = if_else(year < 1920, "pre-1920", "post-1920"))

ggplot(data = runs_scored, aes(x = year, y = runs_scored, colour = period)) +
  geom_point() +
  geom_smooth(color = "black") +
  labs(title = "Runs Scored Through The Years 1903 - 1940",
       x = "Year",
       y = "Number of Runs Scored",
       colour = "Time Period") +
  scale_color_brewer(palette = "Set1") +
  theme_minimal()
```

*In the graph, runs begin to increase during the 1910s. After 1920, the slope of the trend line becomes steeper, and overall run scoring increases until eventually leveling off. This pattern highlights the offensive boost that resulted from banning the spitball.*

**Rule Change #2: The Introduction of a Designated Hitter**

*One pivotal change occurred in 1973, when the American League introduced the designated hitter (DH). Pitchers were generally poor hitters, so to increase offense and make games more exciting for fans, the DH was introduced to bat in place of the pitcher. The National League did not adopt this rule at the time, creating a major difference between the two leagues.*

*Although both leagues continued to field the same number of players, the American League benefited offensively from allowing a stronger hitter to replace the pitcher in the lineup. As a result, the AL recorded higher numbers of runs and home runs compared to the NL.*

*The following boxplots compare the American League and National League before and after 1973 for runs scored and home runs. The analysis uses data from 1903 to 2020. While the National League adopted the DH in 2022, including all prior years helps highlight the long‑term effects of the rule difference. The years 1981, 1994, and 2020 were removed due to shortened seasons caused by strikes and the COVID‑19 pandemic, which could have distorted the results.*

```{r}
designated_hitter <- Baseball |> filter(!year %in% c(1981, 1994, 2020)) |>
  mutate(period = if_else(year < 1973, "pre-1973", "post-1973"))

ggplot(data = designated_hitter, aes(x = league_id, y = runs_scored, fill = period)) +
  geom_boxplot() +
  facet_wrap(~period) +
  labs(title = "Number of Runs Scored Between Each League",
       x = "League",
       y = "Number of Runs",
       fill = "Time Period") +
  theme_minimal()
```
*Before 1973, the AL and NL show similar distributions of runs scored. After 1973, the American League has a noticeably higher distribution of runs than the National League. Additionally, overall run totals are higher in the post‑1973 period, which may also reflect improvements in player ability over time.*

```{r}
ggplot(data = designated_hitter, aes(x = league_id, y = homeruns, fill = period)) +
  geom_boxplot() +
  facet_wrap(~period) +
  labs(title = "Number of Home Runs Between Each League",
       x = "League",
       y = "Number of Homeruns",
       fill = "Time Period") +
  theme_minimal()
```
*The difference between leagues is less pronounced for home runs than for runs scored, but post‑1973 the American League still shows higher home‑run totals than the National League.*

**Rule Change #3: The Strike Zone Change**

*The strike zone is one of the most subjective aspects of baseball. While it is officially defined based on a batter’s height, in practice it is interpreted by umpires and has been redefined multiple times to adjust for changes in the game. In 1963, one of the most significant strike zone changes occurred when the zone was expanded vertically from the batter’s knees to their shoulders, instead of the armpits. This change favored pitchers and led to a increase in strikeouts.*

*By 1968–1969, pitcher dominance had increased so dramatically that the period was nicknamed “The Year of the Pitcher.” In response, MLB implemented another rule change in 1969 that reduced the strike zone back to the armpits and lowered the pitching mound from 15 inches to 10 inches. These changes temporarily slowed the growth of strikeouts, although they did not reverse the long‑term upward trend.* 

```{r}
Pitcherstrikezone <- Baseball |> select(1,6) 

pitcherstrikeouts <- Pitcherstrikezone |> drop_na() |> group_by(year) |> summarise(Total = sum(strikeouts_by_pitchers)) 

library(plotly)

baseballplot <- ggplot(data = pitcherstrikeouts, aes(x = year, y = Total, text = paste("Season:", year,
                                        "Strikeouts:", Total))) +
  geom_point() +
  theme_minimal() +
  geom_vline(xintercept = 1963) +
  geom_vline(xintercept = 1969) +
  labs(title = "Strikeouts by Pitchers per Year",
       x = "Year",
       y = "Number of Strikeouts") +
  lims(y = c(0, NA))
baseballplot

ggplotly(baseballplot, tooltip = "text")
```

*The interactive Plotly graph shows total pitcher strikeouts by year. The vertical lines indicate the major strike zone changes in 1963 and 1969. A sharp increase in strikeouts is visible during the late 1960s, highlighting the "Year of The Pitcher". Outliers appear in 1981, 1994, and 2020 due to strikes and the COVID‑19 pandemic, which reduced the number of games played.*

*Since 1969, additional strike zone adjustments occurred in 1988 and 1996, yet strikeouts continued to rise. This long‑term increase may be due to higher pitch velocity, pitcher specialization, and strategic bullpen usage. In 2026, MLB introduced the Automated Ball‑Strike (ABS) system, formally defining the strike zone as 27% to 53.5% of a player’s listed height, representing a major structural shift toward a standardized, percentage‑based strike zone.*

**More Analysis**

*The following graph is a Shiny App in which the user can select the team of interest and a variable of interest. This app uses the full dataset with all the years and all numerical variables. The graph will only display the years the baseball team existed. The years will end at 2020 since that is the max year in the dataset. Here you can view how teams improved or declined in skill over the years. Use can also use the app to see how the rule changes affected individual teams separately over the years.*

```{r}
library(shiny)
library(tidyverse)
library(lubridate)
library(glue)

mlb_teams2 <- mlb_teams |> 
  select(1, 4:8, 13:39, 41) #Numerical variables only

team_choices <- mlb_teams2 |>
  distinct(team_name) |>
  arrange(team_name) |>
  pull(team_name)

numeric_vars <- mlb_teams2 |>
  select(where(is.numeric)) |>
  select(-year) |>
  names()

ui <- fluidPage(
  titlePanel("MLB Teams Statistics Throughout History"),

  sidebarLayout(
    sidebarPanel(
      selectInput(
        inputId = "team_name",
        label = "Choose a Team:",
        choices = team_choices,
        selected = "Anaheim Angels"
      ),

      selectInput(
        inputId = "yvar",
        label = "Choose a Variable:",
        choices = numeric_vars,
        selected = numeric_vars[1]
      )
    ),

    mainPanel(
      plotOutput("baseball_plot")
    )
  )
)

server <- function(input, output) {

  output$baseball_plot <- renderPlot({

    df_team <- mlb_teams2 |>
      filter(team_name == input$team_name)

    ggplot(df_team, aes(x = year, y = .data[[input$yvar]])) +
      geom_line(linewidth = 1) +
      theme_minimal(base_size = 14) +
      labs(
        title = glue("{input$team_name} Statistics: {input$yvar}"),
        x = "Year",
        y = str_replace_all(input$yvar, "_", " ") |> str_to_title()
      )

  })
}

shinyApp(ui, server)
```

**Conclusion**

*One limitation of this analysis is that many factors beyond rule changes may have influenced the observed trends. Baseball has continually evolved, and confounding variables such as new technology, player demographics, more rule changes, and historical events likely affected the data.*

*For future analysis, I would explore how the implementation of ABS affects win‑loss outcomes and consistency in performance. Additionally, it would be interesting to examine how historical events such as World War I and World War II impacted baseball performance and league statistics.*
