---
title: "334 Final Project"
format: revealjs
editor: visual
---

```{r}
library(readr)
library(ggplot2)
library(tidyverse)
library(dplyr)
mlb_teams <- read_csv("data-finalProject/mlb_teams.csv")
Baseball <- mlb_teams |> select(1:2, 13, 18, 20, 21, 35) |> slice(261:2784)

stolen_bases <- Baseball |> filter(year <= "1940") |>
mutate(period = if_else(year < 1920, "pre-1920", "post-1920"))

designated_hitter <- Baseball |> filter(year >= "1961") |>
  mutate(period = if_else(year < 1973, "pre-1973", "post-1973"))

Pitcherstrikezone <- Baseball |> select(1,7) 

pitcherstrikeouts <- Pitcherstrikezone |> drop_na() |> group_by(year) |> summarise(Total = sum(strikeouts_by_pitchers)) 
```

## Data Introduction

 - Baseball has been a professional sport in America for 150 years and has experienced MANY rule changes either to improve the game or define the game more, these rules changes made a lasting impact on how players played baseball

## Data Set
 - The mlb_teams dataset used it from openintro linked here: https://www.openintro.org/data/index.php?data=mlb_teams)  - looks at every professional team that exists or has existed from 1876 to 2020 with various numerical and categorical variables on each team's performance. 

 - I sliced the dataset to only include world series years (1903-2020) and selected variables of interest: year, league_id, runs_scored, homeruns, stolen_bases and strikeouts_by_pitchers. 

## Question of Interest
 - How did some of the biggest rule changes in the history of baseball affect the game:
 
    1) Abolition of Spitball (1920)
    2) Introduction of a Designated Hitter (1973)
    3) Strike zone shifts & mound reduction (1963 and 1969)

## Rule Change #1

Abolition of spitball led to a decrease in stolen bases and a increases in runs scored per year
```{r}

ggplot(data = stolen_bases, aes(x = year, y = stolen_bases, colour = period)) +
  geom_point() +
  geom_smooth(color = "black") +
  labs(title = "Stolen Bases Through The Years 1903 - 1940",
       x = "Year",
       y = "Number of Stolen Bases",
       colour = "Time Period") +
  scale_color_brewer(palette = "Set2") +
  theme_minimal(base_size = 24)
```

## Rule Change #1b

```{r}
runs_scored <- Baseball |> filter(year <= "1940") |>
  mutate(period = if_else(year < 1920, "pre-1920", "post-1920"))

ggplot(data = runs_scored, aes(x = year, y = runs_scored, colour = period)) +
  geom_point() +
  geom_smooth(color = "black") +
  labs(title = "Runs Scored Through The Years 1903 - 1940",
       x = "Year",
       y = "Number of Runs Scored",
       colour = "Time Period") +
  scale_color_brewer(palette = "Set1") +
  theme_minimal(base_size = 24)
```

## Rule Change #2

Introduction of a Designated Hitter led to the AL league to have more runs scored and home runs than the NL
```{r}
ggplot(data = designated_hitter, aes(x = league_id, y = runs_scored, fill = period)) +
  geom_boxplot() +
  facet_wrap(~period) +
  labs(title = "Number of Runs Scored Between Each League",
       x = "League",
       y = "Number of Runs",
       fill = "Time Period") +
  theme_minimal(base_size = 24)
```

## Rule Change #2b

```{r}
ggplot(data = designated_hitter, aes(x = league_id, y = homeruns, fill = period)) +
  geom_boxplot() +
  facet_wrap(~period) +
  labs(title = "Number of Homeruns Scored Between Each League",
       x = "League",
       y = "Number of Homeruns",
       fill = "Time Period") +
  theme_minimal(base_size = 24)
```

## Rule Change #3 

Strike zone change in 1963 to favor pitchers by that led to "Year of the Pitcher" 1968-1969 so it change again in 1969 with the mound reducing form 15 inches to 10 inches to help batters more

```{r}
library(plotly)

baseballplot <- ggplot(data = pitcherstrikeouts, aes(x = year, y = Total, text = paste("Season:", year,
                                        "Strikeouts:", Total))) +
  geom_point() +
  theme_minimal() +
  geom_vline(xintercept = 1963) +
  geom_vline(xintercept = 1969) +
  labs(title = "Strikeouts by Pitchers per Year",
       x = "Year",
       y = "Number of Strikeouts") +
  lims(y = c(0, NA))
baseballplot

ggplotly(baseballplot, tooltip = "text")
```

## More Applications on Final Project

 - Created a shiny app that allows user to select any team and a variable of interest (numerical only) to see trend throughout that team's existence. (This shiny app uses the whole dataset)
 
## Future Directions

 - It would be interesting to see how ABS (2026) affects the game like number of walks and strikes
 - Go Yankees!
